---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# readepi

*readepi* provides functions for importing epidemiological data into __R__ from
common _health information systems_.

<!-- badges: start -->
[![License: MIT](https://img.shields.io/badge/License-MIT-yellow.svg)](https://opensource.org/licenses/MIT)
[![R-CMD-check](https://github.com/epiverse-trace/readepi/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/epiverse-trace/readepi/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/epiverse-trace/readepi/branch/main/graph/badge.svg)](https://app.codecov.io/gh/epiverse-trace/readepi?branch=main)
[![lifecycle-concept](https://raw.githubusercontent.com/reconverse/reconverse.github.io/master/images/badge-concept.svg)](https://www.reconverse.org/lifecycle.html#concept) 
<!-- badges: end -->


## Installation

You can install the development version of _readepi_ from
[GitHub](https://github.com/) with:

``` r
# install.packages("remotes")
remotes::install_github("epiverse-trace/readepi")
```

## Example

These examples illustrate some of the current functionalities:

```{r example}
library(readepi)

# example of read_stuff():
# (a placeholder function returning its own argument)
path_to_file <- "some_path_here"
read_stuff(path_to_file)

```





## Development

### Lifecycle

This package is currently a *concept*, as defined by the [RECON software
lifecycle](https://www.reconverse.org/lifecycle.html). This means that essential
features and mechanisms are still being developed, and the package is not ready
for use outside of the development team.


### Contributions

Contributions are welcome via [pull requests](https://github.com/epiverse-trace/readepi/pulls).

Contributors to the project include:

* Thibaut Jombart (author)
* Joshua W. Lambert (contributor)

### Code of Conduct
Please note that the linelist project is released with a 
[Contributor Code of Conduct](https://contributor-covenant.org/version/2/0/CODE_OF_CONDUCT.html).
By contributing to this project, you agree to abide by its terms.

### Command line instructions: Create a new Github repository using a template

There's a shortcut to this, which is provided here for people short on time, but the scenic route follows.

**Github command line client**

Install the Github command line client `gh` following the instructions for your OS here: https://github.com/cli/cli.
Run `gh auth login` to link your device with your Github account the first time you use the CLI.

Once installed and authenticated, run:

```sh
gh repo create epiverse-trace/mypackage --public --template=epiverse-trace/packagetemplate
```

Clone the repository as usual using your preferred method.

### Command line instructions: Create a repo and R package from scratch

#### Make an R project

```sh
Rscript -e 'devtools::create("mypackage")'
```

or

```sh
Rscript -e 'usethis::create_package("mypackage")'
```

#### Initialise a git repository

```sh
cd mypackage
git init
```

#### Create a remote Github repository

**Github command line client**

Install the Github command line client `gh` following the instructions for your OS here: https://github.com/cli/cli.
Run `gh auth login` to link your device with your Github account the first time you use the CLI.

Create the package repository in the _Epiverse TRACE_ organisation.

```sh
# still in mypackage
gh repo create epiverse-trace/mypackage
```

#### Initial commit to main

```sh
git push --set-upstream origin main
```

#### Get R-related .gitignore

Getting `R.gitignore` from https://github.com/github/gitignore, and renaming it to `.gitignore`.

`wget` is cross platform and should be available on most systems. Alternatives using `curl` exist.

```sh
wget https://raw.githubusercontent.com/github/gitignore/main/R.gitignore
mv R.gitignore .gitignore
```

#### Set up R package components

These next steps are from the R command line.
If you are prompted to copy or edit text in the terminal (especially if you launched R from the terminal using `$ R`), you might be in Vim. Exit by typing `:wq`, and edit these files in a text editor.

1. Populate fields in the DESCRIPTION and NAMESPACE

```r
# e.g. using data.table
usethis::use_package("data.table")
```

Actually `data.table` requires a bit more configuration to use many of its features, such as `:=` and `.SD`. Use instead:

```r
usethis::use_data_table()
```

2. Set up unit testing infrastructure

```r
usethis::use_testthat()

# and a basic test file
usethis::use_test("basic-test")

# code coverage YAML for codecov
usethis::use_coverage("codecov")
```

3. Set a licence

We prefer the MIT licence, primarily because it is short and easy to read.

```r
usethis::use_mit_licence()
```

4. Make a README page

```r
usethis::use_readme_rmd()
```

Readme.md can be updated manually from the Readme.Rmd created above, using

```r
devtools::build_readme()
```

or using a Github Actions setup (see the examples here https://github.com/r-lib/actions/tree/v2-branch/examples).

5. Set up continuous integration using Github Actions

```r
# to check package installation
usethis::use_github_actions_check_standard()

# to report code coverage results to codecov
usethis::use_github_action("test-coverage")
```

6. Set up documentation

a. Document the package as required using

```r
devtools::document()
```

b. Set up a `pkgdown` website.

```r
# configure the pkgdown YAML.
# can be saved and edited later
usethis::use_pkgdown()

# build the website using
pkgdown::build_site()
```

Or set up a Github Actions job using `usethis::use_github_action("pkgdown")`.
