---
title: "Custom analysis"
output:
  html_document:
    df_print: paged
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(sivirep)
```

Below is a basic set of instructions for using `sivirep` in the following 
scenarios:

- You have already created an `.Rmd` file and want to edit a report.
- You want to perform custom analyses without using an `.Rmd` file.

### Setting

You can start by importing the package using the following command:

```{r import-sivirep}
library(sivirep)
```

### 1. Importing sivigila data

The SIVIGILA source provides historical data up to the most recent closed 
epidemiological year.  The closing of an epidemiological year typically occurs 
in April of the following year.For example, if you are using `sivirep` in March 
2025, you will likely have access to historical data up to December 2024  for 
most diseases, with a few exceptions.

Before using the function, please check the available diseases and years with 
the following command:

```{r list-events, results = 'hide', eval = FALSE}
list_diseases <- list_events()
```

Once you have decided the disease and year you want, you can use
`import_data_event` to import the disease data from SIVIGILA using the disease 
or event name and year (you can download data for multiple years).

```{r import-data}
disease_data <- import_data_event(nombre_event = "dengue",
                                  years = 2020,
                                  ruta_dir = tempdir())
```

>💡 **Tip - Avoid time delays when importing data** <br/>
   `sivirep` is meant to assist with access to SIVIGILA source.
    The process of downloading disease information may take a few minutes 
    depending on the size of the dataset. To avoid re-downloading the same data 
    each time, you can use the `cache = TRUE` parameter in the function 
    `import_data_event`.

>💡 **Tip - Download disease data for multiple years** <br/>
    With the function `import_data_event`, it is possible to download data for 
    multiple years. For example, if you want to download data for 3 years of a 
    particular diease, you can use the `year` parameter as follows:
   `import_data_event(data_event = "dengue", years = c(2021, 2019, 2018), cache = TRUE)`
    You can also specify non-consecutive years:
   `import_data_event(data_event = "dengue", years = c(2024, 2018, 2013), cache = TRUE)`

### 2. Cleaning sivigila data

SIVILA data is a highly trusted official source of information with ISO
certification of data quality. Nevertheless, some potential outliers or 
anomalies may be present in the data, requiring additional cleaning.

`sivirep` provides a wrapper generic function called
`limpiar_data_sivigila`, which helps identifying and correcting errors, 
inconsistencies, and discrepancies to improve the quality and accuracy of 
datasets. This process may include tasks such as removing duplicates, correcting typographical errors, imputing missing values, and validating entries. It can 
also involve cleaning improbable dates, standardizing geolocation codes, 
harmonizing column names and age categories, all in accordance with the disease 
protocols defined by the INS.

```{r clean-data}
cleaned_disease_data <- limpiar_data_sivigila(data_event = disease_data)
```

These may include internal functions such as: 

-   `limpiar_edad_event`: function that cleans and standardises the column names 
    of SIVIGILA data.
-   `limpiar_edad_event`: function that cleans and standardizes ages from 
    disease data to years, according to INS clasification.
-   `limpiar_val_atipic`: function  that cleans outliers from disease data.
-   `limpiar_fecha_event`: function that cleans and standardizes dates from 
    disease data.
-   `estandarizar_geo_cods`: function that standardizes the geographical codes 
    of the disease data, according to DIVIPOLA codification.
-   `convert_edad`: function that converts  ages to years according to SIVIGILA 
    measurement units.

You may want to use these functions individually or simply use the
generic wrapper cleaning function.


### 3. Filter cases

`sivirep` provides a function called `geo_filtro`, which you can use to filter 
disease data by department or municipality . It allows you to create a 
subnational report.

```{r filter-data}
filtered_disease_data <- geo_filtro(data_event = cleaned_disease_data,
                                    "Choco")
``` 

### 4. Temporal distribution of cases

In `sivirep` the temporal distribution of cases is defined by the two variables: 
symptom onset date and notification date. For each of these
variables, there are specialised functions to group the data and generate
the plots.

#### 4.1. Grouping data by symptom onset at the desired temporal scale

To generate the distribution of cases by symptom onset date, it is
necessary to group the data by this variable. `sivirep` provides a
function called `group_onset_symptoms` for this task.

```{r group-symptom-onset}
cases_onset_symptoms <- agrupar_fecha_inisintomas(data_event =
                                                    cleaned_disease_data)
```

>💡 **Tip - Get the first n months with most cases** <br/>
    When building a report section or analysing this data it can be
    useful to get the months with the most cases. In `sivirep` you can
    use the function `obtener_meses_mas_casos` to get this information.

The graph that allows to visualise this distribution must be generated
with the function `plot_fecha_inisintomas`. You can customise the plot's ticks 
by day (`dia`), month (`mes`) or epidemiological week (`semanaepi`), as shown 
below:

```{r plot-symptom-onset, fig.height = 5, fig.width = 6}
plot_fecha_inisintomas(data_agrupada = cases_onset_symptoms,
                       uni_marca = "semanaepi")
```


### 5. Age and sex

### 5.1. Sex variable

When analysing or reporting disease data, it is often necessary to
determine the distribution of cases by gender or sex. However, the
SIVIGILA source only records sex.

`sivirep` provides a function that groups and computes percentages
by sex automatically after the cleaning process.

```{r group-sex}
cases_sex <- agrupar_sex(data_event = cleaned_disease_data,
                         porcentaje = TRUE)
```

Also, `sivirep` provides a function called `plot_sex`to plot this distribution:

```{r plot-sex, fig.height = 5, fig.width = 7}
plot_sex(data_agrupada = cases_sex)
```

To generate the distribution of cases by sex and epidemiological week, you can use
the `agrupar_sex_semanaepi` function provided by `sivirep`:

```{r group-semanaepi}
cases_sex_epiweek <- agrupar_sex_semanaepi(data_event = cleaned_disease_data)
```

The corresponding visualisation function that provides `sivirep` is `plot_sex_semanaepi`:

```{r plot-semanaepi, fig.height = 5, fig.width = 7}
plot_sex_semanaepi(data_agrupada = cases_sex_epiweek)
```

### 5.2. Age variable

Age is an important variable to analyse, as it is a known risk factor
for many diseases. Certain diseases and conditions tend to occur more
frequently in specific age groups, and this distribution can help
identify populations at higher risk and implement targeted prevention
and control strategies.

`sivirep` provides a function called `agrupar_edad`, which can group
disease data into age groups. By default, this function produces age ranges
with intervals of 10 years. Users can  customise different age ranges.

```{r group-age}
cases_age <- agrupar_edad(data_event =  cleaned_disease_data,
                          interval_edad = 10)
```

The corresponding plot function is `plot_edad`:

```{r plot-age, fig.height = 5, fig.width = 6}
plot_edad(data_agrupada = cases_age)
```

### 5.3. Age and sex simultaneously

`sivirep` provides a function called `agrupar_edad_sex`, which can group
disease data by age ranges and sex simultaneously and calculate the 
number of cases and their percentages. Also, the age interval can be
customised.

```{r group-age-sex}
cases_age_sex <- agrupar_edad_sex(data_event = cleaned_disease_data,
                                  interval_edad = 10,
                                  porcentaje = TRUE)
```

The function that `sivirep` provides to plot this corresponding distribution is `plot_edad_sex`:

```{r plot-age-sex, fig.height = 5, fig.width = 6}
plot_edad_sex(data_agrupada = cases_age_sex)
```

### 6. Spatial distribution of cases

Obtaining the spatial distribution of cases is helpful to identify areas
with a high concentration of cases, disease clusters and environmental
or social risk factors.

In Colombia, there are 32 administrative geographic units (adm1) called
departments. 

`sivirep` provides a function called `agrupar_mpio` which
allows you to obtain disease data grouped by municipalities of a specific 
department — in other words, the department's case distribution.

```{r group-mun}
spatial_mun_dist <- agrupar_mpio(data_event =  filtered_disease_data,
                                 dpto = "Choco")
```

There is also a function to obtain the case distribution across all departments 
of Colombia, called `agrupar_dpto`:

```{r group-dpto}
spatial_dept_dist <- agrupar_dpto(data_event =  cleaned_disease_data)
```

Currently, with the function called `plot_map` the user can produce
a static map of Colombia with the distribution of cases or incideces by departments and municipalities.

>💡 **Tip - Avoid delays when generating the map** <br/>
 It is necessary to download the shapefiles of Colombia to generate the map. 
 To avoid re-downloading them each time, you can use the `cache = TRUE` 
 parameter in the `plot_map` function.

```{r plot-map, fig.height = 7, fig.width = 7}
map <- plot_map(data_agrupada = spatial_mun_dist,
                col_distribucion = "casos",
                ruta_dir = tempdir())
map
```

> 💡 **Tip - Get the row with most cases** <br/>
  When building a report section or analysing this data it can be
  helpful to know which sex, age, etc., has the most cases, in `sivirep` you
  can use the function `obtener_fila_mas_casos` to get this information. This
  function works with any dataset that contains a column called
  `"cases"` by any level of aggregation.

### 7. Incidence

With `sivirep`, you can calculate incidence rates by geography or sex using 
DANE population projections or population-at-risk depending on event and data 
availability.

> 📑 **Note:** sivirep` does not include population-at-risk data for all events. 
  However if you have this information, you can provide it using the 
  `data_incidencia` parameter in each function.

#### 7.1 Incidence by geography

You can calculate incidence by departments or municipalities of a specific
department using the `calcular_incidencia_geo` function:

```{r incidence-muns, echo = FALSE, error = FALSE, warning = FALSE, include = TRUE, message = FALSE}
incidence_muns <-
     calcular_incidencia_geo(data_agrupada = spatial_mun_dist,
                             ruta_dir = tempdir())
```

##### 7.1.1 Map of incidence by geography

With the function `plot_map` you can visualize and generate a map with the 
incidence for a specific deparment or municipality, depending on your selected 
parameters and calculations.

```{r plot-map-incidence, fig.height = 7, fig.width = 7}
map <- plot_map(data_agrupada = incidence_muns$data_incidencia,
         ruta_dir = tempdir())
map
```

#### 7.2 Incidence by sex

`sivirep` provides a function called `calcular_incidencia_sex` that 
allows you to calculate incidence by sex at national level (Colombia), or for a 
specific deparment or municipality.

```{r incidence-sex, echo = FALSE, error = FALSE, warning = FALSE, include = TRUE, message = FALSE}
incidence_sex <-  calcular_incidencia_sex(
   data_agrupada = cases_sex,
   dpto = "Choco",
   ruta_dir = tempdir()
 )
```

#### 7.3 Total incidence in the country

The `calcular_incidencia` function allows you to  calculate incidence for the 
entire country:

```{r total-incidence, echo = FALSE, error = FALSE, warning = FALSE, include = TRUE, message = FALSE }
total_incidence <-
     calcular_incidencia(data_agrupada = cleaned_disease_data,
                             ruta_dir = tempdir())
```
