---
title: "Customized Analyses"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Below is a basic set of instructions for using `sivirep` if:

* You have already produced an .Rmd file and wish to edit a report.
* You wish to perform customized analyses without an .Rmd file.

### Setup

You can start by importing the package with the following command:

```{r import-sivirep}
library(sivirep)
```

### 1. Importing Data from SIVIGILA

The SIVIGILA source provides historical case list data up to the last closed epidemiological year. For most diseases, the closing usually happens in April of the following year.

You can check the available diseases and years using:

```{r eventos, results = 'hide'}
event_list <- list_events()
```

Once you have decided on the disease and year, use `import_data_event` to import data from SIVIGILA using a parameterized format based on disease and year:

```{r import-data-event}
data_event <- import_data_event(
  year = 2020,
  nombre_event = "Dengue",
  cache = TRUE
)
```

> 💡 **Tip**: Avoid delays when importing data by setting `cache = TRUE`.

### 2. Cleaning SIVIGILA Data

SIVIGILA data is a highly reliable official information source, with ISO certification for data quality. However, there may sometimes be outliers in the data that require additional cleaning.

`sivirep` provides a generic function called `limpiar_data_sivigila` that encompasses various tasks to identify and correct errors, inconsistencies, and discrepancies in datasets in order to improve their quality and accuracy. This process may include the removal of duplicates, correction of typographical errors, replacement of missing values, and data validation, among other tasks such as eliminating improbable dates, cleaning geolocation codes, and standardizing column names and age categories.

```{r limpiar-data-sivigila}
data_event <- limpiar_data_sivigila(data_event = data_event)
```

The cleaning functions within `limpiar_data_sivigila` have been compiled and created based on the experience of field epidemiologists. These may include internal functions such as:

- `limpiar_encabezado`: a function that cleans and standardizes column names in SIVIGILA case list data based on the SIVIGILA data dictionary.

- `limpiar_edad_event`: a function that cleans the ages in SIVIGILA case list data.

- `limpiar_fecha_event`: a function that cleans the dates in disease data.

The user can use these functions individually or simply use the generic wrapper function `limpiar_data_sivigila`.

### 3. Filtering Cases

`sivirep` provides a function that allows filtering disease data by department or municipality name called `geo_filtro`. This enables the user to create a subnational-level report by selecting specific cases based on geographic location.

```{r geo-filtro}
filtered_data_event <- geo_filtro(
  data_event = data_event,
  dpto = "Choco"
)
```
### 4. Temporal Distribution of Cases

In `sivirep`, the temporal distribution of cases is defined by the variables "symptom onset date" and "notification date." For each of these variables, there are specialized functions to group the data and generate graphs.

#### 4.1. Grouping by Symptom Onset Date

To generate the distribution of cases by symptom onset date, it is necessary to group the data by these variables. `sivirep` provides a function that allows this grouping called `agrupar_fecha_inisintomas`.

```{r agrupar-fecha-inicio-sintomas}
cases_ini_symptoms <- agrupar_fecha_inisintomas(
  data_event =
    data_event
)
```

> 💡 **Tip**: Get the top n months with the most cases  
>
> When building a section of the report or analyzing these data, it may be helpful to find the months with the most cases. In `sivirep`, you can use the function `obtener_meses_mas_casos` to obtain this information.

The graph to visualize this distribution should be generated with the `plot_fecha_inisintomas` function. Keep in mind that even if you have grouped the data by day, you might prefer to represent it by epidemiological week, as follows:

```{r plot-fecha-inicio-sintomas, fig.height = 4, fig.width = 8}
plot_fecha_inisintomas(
  data_agrupada = cases_ini_symptoms,
  uni_marca = "semanaepi"
)
```

### 5. Age and Sex Analysis

#### 5.1. Sex Variable

When analyzing or reporting disease data, it is often necessary to determine the distribution of cases by gender or sex. However, the SIVIGILA source only records sex.

`sivirep` provides a function that automatically aggregates and calculates percentages by sex after the cleaning process.

```{r agrupar-sexo}
cases_sex <- agrupar_sex(
  data_event = data_event,
  porcentaje = TRUE
)
```

Additionally, `sivirep` includes a function to generate the graph for this variable called `plot_sex`:

```{r plot-sexo, fig.height = 5, fig.width = 7}
plot_sex(data_agrupada = cases_sex)
```

The distribution of cases by sex and epidemiological week can be generated using the function `agrupar_sex_semanaepi` provided by `sivirep`.

```{r agrupar-sex-semana-epidemiologica}
cases_sex_week_epi <- agrupar_sex_semanaepi(data_event = data_event)
```

The corresponding visualization function is `plot_sex_semanaepi`, which `sivirep` provides to show the distribution of cases by sex and epidemiological week.

```{r plot-sex-semana-epidemiologica, fig.height = 7, fig.width = 8}
plot_sex_semanaepi(data_agrupada = cases_sex_week_epi)
```

#### 5.2. Age Variable

Age is an important variable to analyze, as it is a known risk factor for many diseases. Certain diseases and conditions tend to occur more frequently in specific age groups, and this distribution can help identify populations at higher risk and implement targeted prevention and control strategies.

`sivirep` provides a function called `agrupar_edad`, which can group disease data by age groups. By default, this function produces age ranges in 10-year intervals. Additionally, users can customize a different age range.

```{r agrupar-edad}
cases_age <- agrupar_edad(data_event = data_event, interval_edad = 10)
```

The corresponding visualization function is `plot_edad`.

```{r plot-edad, fig.height = 3, fig.width = 6}
plot_edad(data_agrupada = cases_age)
```

#### 5.3. Age and Sex Together

`sivirep` proporciona una función llamada `agrupar_edad_sex`, que puede
agrupar los datos de enfermedades por rangos de edad y sexo de forma
simultánea y obtener el número de casos y los porcentajes
correspondientes. Además, permite personalizar el intervalo de edad.

```{r agrupar-edad-sexo}
cases_age_sex <- agrupar_edad_sex(
  data_event = data_event,
  interval_edad = 10
)
```

La función de visualización correspondiente es `plot_edad_sex`.

```{r plot-edad-sexo, fig.height = 3, fig.width = 6}
plot_edad_sex(data_agrupada = cases_age_sex)
```

### 6. Spatial Distribution of Cases

Obtaining the spatial distribution of cases is useful for identifying areas with a high concentration of cases, disease clusters, and environmental or social risk factors.

In Colombia, there are 32 administrative geographic units (adm1) called departments. `sivirep` provides a function called `agrupar_mpio` that allows obtaining a data.frame of cases grouped by department or municipality.

```{r agrupar-municipios}
dist_esp_dept <- agrupar_mpio(
  data_event = filtered_data_event,
  dpto = "Choco"
)
```

Currently, with the function called `plot_map`, the user can generate a static map of Colombia showing the distribution of cases by departments and municipalities.

```{r plot-mapa, results = 'hide', echo = TRUE, error = FALSE, warning = FALSE, include = FALSE, message = FALSE}
mapa <- plot_map(
  data_agrupada = dist_esp_dept, dpto = "Choco",
  col_distribucion = "casos",
  cache = TRUE
)
```

```{r mapa, fig.height = 14, fig.width = 10}
mapa
```

> 💡 **Tip**: Obtain the row with the highest number of cases  
>
> When building a section of the report or analyzing these data, it may be useful to know which variable has the most cases. In `sivirep`, you can use the function `obtener_fila_mas_casos` to get this information. This function works with any dataset that contains a column named `"casos"` at any level of aggregation.