---
title: "Get started"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(sivirep)
```

## Description

Current version of *sivirep* provides functions for data wrangling and
automated reports based on data from SIVIGILA, which is the official 
epidemiological surveillance system of Colombia, South America.

## Disclaimer

The use of this package, as well as the data, generated reports, and other 
products derived from it, is the sole responsibility of the user. Neither the 
authors of the package, nor the Pontificia Universidad Javeriana, nor the 
source of the information assume any responsibility for the results obtained 
or the use made of these products.

## Motivation

Latin America has progressed in the quality of epidemiological
notification and surveillance systems. Particularly, Colombia has
improved over the years the quality and openness of its official
epidemiological surveillance system, SIVIGILA. This system is regulated 
by Colombia's National Institute of Health and operated by thousands of health 
workers at local secretaries of health, hospitals, and local notification units.

However, some challenges remain particularly at local levels in terms of
timeliness and quality of epidemiological analytics and epidemiological
reports. These tasks may involve a great effor of manual labor reinforced
by limitations in training for data analytics, time, technology and
quality of internet access in some locations.

`sivirep` is aimed at providing a set of customisable functions for:

1)  Downloading, pre-processing and preparing SIVIGILA data for further
    analysis.
2)  Producing customisable epidemiological automated reports.
3)  Getting feedback on the surveillance system to the source provider.

## Potential users

-   Public health professionals and field epidemiologist users of
    SIVIGILA source at local levels
-   Epidemiology and public health students
-   National and international researchers and data analysts

## Future versions

Future versions of `sivirep` may include:

-   Interaction with other data sources in Colombia
-   Other epidemiological surveillance systems in Latin America

## Contributions

Contributions are welcome via [pull
requests](https://github.com/epiverse-trace/sivirep/pulls).

Contributors to the project include:

**Autores**: [Geraldine Gómez-Millán](https://github.com/GeraldineGomez), 
[Zulma M. Cucunubá](https://github.com/zmcucunuba), Jennifer A. Mendez-Romero y [Claudia Huguett-Aragón](https://github.com/chuguett)

**Contribuyentes**:
[Hugo Gruson](https://github.com/Bisaloo), [Juanita Romero-Garcés](https://github.com/juanitaromerog), [Andrés Moreno](https://github.com/andresmore), 
[Miguel Gámez](https://github.com/megamezl), [Laura Gómez-Bermeo](https://github.com/lgbermeo), Johan Calderón, Lady Flórez-Tapiero, 
Verónica Tangarife-Arredondo y Gerard Alarcon

## Code of Conduct

Please note that the linelist project is released with a [Contributor
Code of 
Conduct](https://contributor-covenant.org/version/2/0/CODE_OF_CONDUCT.html).
By contributing to this project, you agree to abide by its terms.

## Installation

You can install the development version of `sivirep` from GitHub with:

```r
install.packages("pak")
pak::pak("epiverse-trace/sivirep")
```

## Quick start

You can start by importing the package after installation is complete using the 
following command:

```{r import-sivirep, eval = FALSE}
library(sivirep)
```

You can check the diseases and years available for free download using the 
following command:

```{r list-events-ms, eval = FALSE}
list_events <- list_events()
knitr::kable(list_events)
```

`r paste0("<blockquote style=", '"color: #333; cursor:pointer;"', "><details><summary>🦠<b>Listado de enfermedades (haz clic para ver)</b></summary><br>")`

```{r list-events-rs, echo = FALSE}
list_events <- NULL
tryCatch({
  list_events <- list_events()
  knitr::kable(list_events,
             col.names = c("Code", "Disease", "Year"),
             row.names = FALSE, align = "l")
}, error = function(e) {
  list_events <- NULL
})
```

`r if (is.null(list_events)) { paste0("<ul><li><a class=", '"external-link"',  " href=", '"https://epiverse-trace.github.io/sivirep/articles/resources.html#enfermedades-y-a%C3%B1os-disponibles-para-su-descarga"', ">Enfermedades</a></li></ul>")}`

`r paste0("</details></blockquote>")`

## Automated report

Currently, `sivirep` provides a report template called  
`Reporte Evento {sivirep}`, which takes the following input parameters:

-   Name of the disease  
-   Year  
-   Country  
-   Name of the department *(optional)*  
-   Name of the municipality *(optional)*
-   Cache data *(checkbox - optional)* : choose whether the downloaded data for 
    the event and incidence should be stored in the user's session cache to 
    avoid re-downloading it when generating the report again with the same 
    inputs.

These parameters are used to download data from the SIVIGILA source.

To use the report template, you can follow these steps:

> 🎥 [How to generate a report with sivirep?](https://youtu.be/wsgXQKEeg8I)

The report generated using the `sivirep` template looks like this:

> 🎥 [sivirep report](https://youtu.be/NRUNwVrs4io)

If you want to generate the report in PDF format, you need to install LaTeX.  
You can install it by following the instructions in the [R Markdown Cookbook](https://bookdown.org/yihui/rmarkdown-cookbook/install-latex.html).

## Custom analysis

`sivirep` offers a suite of functions that can be used in different scenarios —  
from downloading data to creating custom analyses or building full 
pipelines.

To explore the available functions and examples of their suage, please visit the 
[Custom analysis](https://epiverse-trace.github.io/sivirep/articles/custom_analysis_en.html)page.
